---
title: "Error in age-sex Data of WB"
author: |
      | Saikat Roy
      | IIPS/MBD/2020-22
date: "23/12/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

The errors in the age-sex data of West Bengal as enumerated in the census of West Bengal, 2011 are discussed here.

```{r import, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
library(readxl)
wbagesexdata <- read_excel("DDW-1900C-13.xls", 
    range = "E9:H109", col_names = FALSE)

colnames(wbagesexdata) <- c("age","total","male","female")
wbagesexdata = as.data.frame(wbagesexdata)
wbagesexdata$age = as.character(wbagesexdata$age)
wbagesexdata$age = factor(wbagesexdata$age, levels = unique(wbagesexdata$age))
```

### Single-year Age Groups

Line curves are plotted for total, male and female populations according as the ages.

```{r linecurve}
library(ggplot2)
library(reshape2)
mdf <- melt(wbagesexdata, id.vars="age")


linefig <- ggplot(mdf, aes( x=age, y=value, colour=variable, group=variable )) + 
  geom_line() +
  geom_point() +
  scale_color_manual(values=c("total"="black","male"="red","female"="orange")) +
  scale_linetype_manual(values=c("total"="dashed", "male"="solid","female"="solid")) +
  scale_x_discrete(breaks = wbagesexdata$age[c(T,F,F,F,F)]) +
  ggtitle("Age-Sex Population Distribution, \n West Bengal (2011)") +
  xlab("Population") + ylab("Age (lbd)") +
  theme(plot.title = element_text(hjust = 0.5))

linefig

```

Single year population pyramid with absolute population size values for each age-sex is constructed.

```{r pyramid, fig.height=10, fig.width=12}
malefemale <- wbagesexdata[,c(3,4,1)]
library(pyramid)

pyramid(malefemale,Llab="Males", Rlab="Females", Clab="Age (lbd)",main = "Single-year Age Population Pyramid \n West Bengal, 2011", Cstep = 5)
```

It is clear from the above figures that population count unusually spikes hugely at ages ending 0 and 5. Spikes, albeit smaller, are also noticed at even ages. It is clear that people are biased in favour of reporting ages divisible by 5 and slightly lesser biased towards reporting even ages. The tendency to report odd ages is low. This age-heaping causes errors.


